---
title: "City of Toronto's development applications summary between 2008 and 2021"
author: "Yang Wu"
thanks: "code and data are available at: https://github.com/yangg1224/Toronto-development-applications.git/."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: The author downloads Toronto's development applicatioin data set from Open Data Toronto website and summarize all the approved and in progress applications between 2008 and 2021.For governments, these data provides a overview for them to consider the next step urban planning. For individuals, these data make them aware what infrastructure and new projects are around.
output:
  bookdown::pdf_document2:
  toc: FALSE
Contact: Yangg.wu@mail.utoronto.ca
bibliography: references.bib
---
# Introduction 

Due to the aggravation of COVID, the chance of everyone going out is also decreasing. Therefore, there are fewer and fewer opportunities to discover new changes around you. However, The city develops faster than you think. Have you ever thought about what changes are happening around you every day? City of Toronto's application information portal is a web page where you can find information about the new developments in Toronto. If you just move here or If you are an smart investor interested in real estate, these information will be vital reference for you to choose the right place. 

The paper is organized into three chapters. The following chapter will describe the detailed information about the dataset and talk about the data bias. At the end, the last chapter will do the data visualization and exploratory data analysis. 


# Setup workspace and gather data
In this paper, all the data analyses are conducted by R statistical programming language.[@citeR] Through the **opendatatoronto** package, I download the data set for toronto development applications. This package allows users to download the data set in a reproducible way.[@citeO] In terms of data cleaning, I will use **tidyverse** package.[@tidy] Another two data Visualization packages used in the Exploratory Data Analysis are **ggploat2** [@ggplot] and **kableExtra**[@kable]
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(opendatatoronto)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```
# Data description

## Load Data and introduction
The data are available at this link:https://open.toronto.ca/dataset/development-applications/.
Firstly, I load the data and write them into csv file. With the help of **kableExtra**[@kable], let's have a quite look of the raw data. (Table \@ref(tab:loaddata)) I apologize the font is quite small here and you have to zoom in. 
```{r loaddata, fig.cap="raw data set", echo = FALSE,fig.width=18, fig.height=4}
all_data <- 
  opendatatoronto::search_packages("Development Applications") %>% 
  opendatatoronto::list_package_resources()%>% 
  dplyr::filter(name %in% "Development Applications Data") %>% 
  group_split(name) %>% # Don't totally get this
  map_dfr(get_resource, .id = "file")

write_csv(all_data, "raw_data.csv")

# Have a quick look at the data
all_data<-read.csv("raw_data.csv")
  head(all_data)%>%
  select(-DESCRIPTION)%>%
  kableExtra::kbl(caption = "First 6 rows Raw data loaded ") %>%
  kableExtra::kable_styling(latex_options = "scale_down") # use scale_down option to make the font smaller and knit a completed table to PDF
```

This dataset lists all currently active (open) and inactive (closed) Community Planning applications, Committee of Adjustment applications and Toronto Local Appeal Body appeals received by the City between 2008 till present.The data set was published in three formats, which are CSV, Jason and XML.The latest time when the data was refreshed is Jan 27th, 2021. Also, this dataset is refreshed monthly.

## Data bias 
As we can see from the raw data set, it does not give the reasons why some applications are rejected or closed and why some applications are approved. Though this dataset is just a summary of Toronto's development applications, the decision behind the application exist ethical bias. Firstly, urban planners decide what goes where in the city. Obviously, the achievements of these planners are linked to the comments of the local people. So their decisions are often irrational, because they are affected by the mood of the local residents.[@urban]

Another data bias is that currently urban planners would use data driven model to make a best development plan for the city. However, no matter how precises the data they collect, the model cannot understand the history. "Algorithms have difficulty understanding when the architecture in a certain neighborhood is historic to the city." [@urban]

In terms of limitations of the dataset, there are many missing values in hearing_date column. I had planned to calculate a processing time for each application(hearing date - submitted date). Because of the missing values, the result will be not accurate. Besides, Under the status column, it explains the current status of the application. Because the decision of different types of application is made by different parties, there are more than ten "status" in the column which make readers confused. For example, "approved" and "accepted" are the same thing, but it lists in two ways. what is more,the  visualization tool is not compatible with X,Y coordinates in the dateset. 

# Exploratory Data Analysis

## Data cleaning and preparation 
Firstly, i will remove the rows with missing value to improve the accuracy. **Janitor** package is used to clean the column name. [@jan] Then, different types applications are interpreted as following:
 CD - Condominium
 OZ - Official Plan/Rezoning
 PL - Part Lot
 SA - Site Plan Application
 SB - Sub Division
 CO - Consent
 MV - Minor Variance
I will complete the abbreviation so that readers can easily understand. The application submitted date has been separated into year, month and day for further analysis. 
```{r datacleaning, include=FALSE}
development_applications<-
  all_data %>% 
  janitor::clean_names() %>% # It handles problematic variable names and Make the column names easier to type
  # link here:https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#clean-data.frame-names-with-clean_names
  mutate(application_type = 
           case_when(
             application_type == "CD" ~ "Condominium",
             application_type == "OZ" ~ "Official_Plan/Rezoning",
             application_type == "PL" ~ "Part_Lot",
             application_type == "SA" ~ "Site_Plan_Application",
             application_type == "SB" ~ "Sub_Division",
             application_type == "CO" ~ "Consent",
             application_type == "MV" ~ "Minor_Variance")
  ) %>%
  select(date_submitted, application_type, street_name, postal, description, status) %>%
  separate(date_submitted, into =c("year","month","day"), sep="-", remove = FALSE)


```
## Graph
I am interested in the total numbers of approved applications through the year between 2008 and 2021. So i filter the application status and group by the application types. 
Then, the result is shown in the Figure.  (Figure1 \@ref(fi:graph))
```{r graph,fig.cap="raw data set", echo = FALSE,fig.width=10, fig.height=4}
accepted_applications <- 
  development_applications %>% 
  tidyr::drop_na(status) %>% # We only want rows that have data for status
  filter(status %in% c("Accepted", "Approved", "Council Approved","OMB Approved")) %>%# filter the applications only in accepted status
  group_by(application_type) %>% # We want to know the total number of applications by type
  count(application_type) %>%
  rename(Number = n) 


p<-ggplot(accepted_applications,aes(application_type, Number, fill=application_type))+
  geom_bar(stat ="identity" )+
  ylab("numbers")+
  ggtitle("Total approved application (2005-2021)")+
  coord_flip()+
  theme_gray()

p
```

## Table
If i am person who want buy a house in Adelaide street, I probably would like to know more about the surrounding developments. Using **KableExtra**[@kable], the table below shows what are all the applications constructed or will be constructed near this street in 2020. 
(Table \@ref(tab:table2))

```{r table2, tab.cap="ALL applications in ADELAIDE Street", echo = F}
ADELAIDE_street_applications<-
  development_applications %>%
  tidyr::drop_na(status) %>% # We only want rows that have data for status
  filter(street_name %in% "ADELAIDE") 



ADELAIDE_street_applications %>% 
  filter(year %in% c("2020"))  %>%
  select(-year,-month,-day, -street_name,-postal) %>%
  kableExtra::kbl(caption = "ALL applications in ADELAIDE Street (2021)") %>%
  kableExtra::kable_styling(full_width = F,
                            latex_options = "scale_down",
                            position = "center")%>%
  column_spec(1, color = "black")%>%
  column_spec(2, color = "black")%>%
  column_spec(3, color = "black", width="40em", background = "gray")%>%
  column_spec(4, color = "black")

```


# References


